---
title: "How to add and subtruct Dates and Times in R"
author: "Juan Lorenzo Arellano"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Following the R recipe [How to deal with Date and Time in R](dates_times/How-to-deal-with-Dates-and-Times-in-R.md) where we reviewed the base R capabilities to deal with dates and times, we continue exploring how to add and subtract dates and times, also in base R. I plan to cover [**lubridate**](https://lubridate.tidyverse.org/) and [**clock**](https://clock.r-lib.org/) packages in the future.


## Addition of an scalar to a date

It's very common that we need to calculate an end date for a process or service knowing only the starting date and the duration. For this, we need to add the number of **days** (scalar) to a date. Let's take a look:

```{r eval=F}

start_point <- Sys.Date()
duration_days <- 365                # one year
start_point + duration_days         # [1] "2022-04-20"

duration_days <- 15                 # 15 days
start_point + duration_days         # [1] "2021-05-05"

duration_days <- 3 * 30             # 3 months (approx)
start_point + duration_days         # [1] "2021-07-19"

duration_days <- 16 * 7             # 16 weeks
start_point + duration_days         # [1] "2021-08-10"

class(start_point + duration_days)  # [1] "Date"

```

As you have seen above, adding a scalar to date produces another date pointing at the future. If you want to add weeks, months, etc. you need to convert it to days first. That's a bit tricky because three months aren't exactly 90 days. The exact number depends on which months are those. This has to be pre-computed separately, I'm afraid.

```{r}


```

