---
title: "How to deal with Dates and Times in R"
author: "Juan Lorenzo Arellano (Twitter: @jlarellanoh)"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

A common need that exists in every programming language is parsing strings into dates and converting dates into string. This is specifically important for data analysts because we get data from many different sources and countries where the format changes, and we need to read that data into a R as dates. For instance, 31st of January of 2021 can be seen as:

* **01/31/21**: US format
* **January the 31, 2021**: US format
* **31/01/21**: typically European format
* **31.01.21**: typically German format
* **31-01-21**: typically English format
* **31/JAN/21**: variation of English format
* **31 January 2021**: variation of English format

Conversely, in many occasions we to extract specific parts of the date like day of week, calendar week, month name, day name, etc. or produce a long string containing date parts like "Today is Monday of week 23 (2021)" when creating reports or dashboards.

We are going to see how we can do this transformations in **base R** (we will not cover [**lubridate**](https://lubridate.tidyverse.org/) nor the new kid on the block: [**clock**](https://clock.r-lib.org/)

## How to parse a Character String into a Date

### Default date format

A date is data type or class that exists within R. When you need to write a date into a report, document, csv, etc. you covert into a string. Also, to enter a date into the R terminal you produce a string that has to be converted into an R date type.

```{r eval=FALSE}
mydate_chr <- "2021-04-15"     # we create a date as string
class(mydate_chr)              # it's class is character
[1] "character"

mydate_date <- as.Date(mydate_chr)  # we convert it into a date with default format yyyy-mm-dd 
mydate_date                         
[1] "2021-04-15"

class(mydate_date)             # it's class is Date
[1] "Date"
```

Ok, we've cast a string into a Date with default format, but what is the default format?. As per the documentation `?as.Date`, if we do not indicate any format then the parameter *tryFormats* comes into play, whose default value is *c("%Y-%m-%d", "%Y/%m/%d")*. Let's test it!!

```{r eval=FALSE}
mydates <- c("2021/04/01", "2021/04/02", "2021-04-03", "2021.04.04")  # Format with slashes wins "/"
as.Date(mydates)
[1] "2021-04-01" "2021-04-02" NA           NA 

mydates <- c("2021-04-01", "2021/04/02", "2021-04-03", "2021.04.04")  # Format with dashes wins "-"
as.Date(mydates)
[1] "2021-04-01" NA           "2021-04-03" NA  

```
What's happened here? `format` tries all the patterns in `tryFormats` and we it finds one that works then it keeps using that one discarding the others. So I've changed only the format of the first date and it lead the behavior of `as.Date()`.

### Customized date format

What if the default format is not convinient for us. Then we use the `format` argument of `as.Date()`

```{r eval=F}
mydates <- c("01/01/2021", "02/02/2021")  # Changed to dd/mm/yyyy
as.Date(mydates, format = "%d/%m/%Y")     # Works: [1] "2021-01-01" "2021-02-02"

mydates <- c("01/JAN/21", "02/JAN/21")    # Format with slashes wins "/"
as.Date(mydates, format = "%d/%b/%y")     # Works: [1] "2021-01-01" "2021-01-02"

mydates <- c("Thu, 15 of April, 2021")          # Format with slashes wins "/"
as.Date(mydates, format = "%a, %d of %B, %Y")   # Works: [1] "2021-04-15"

```

There are plenty of *conversion specifications* that you can query on the documention of `strptime()` or [here](https://rdrr.io/r/base/strptime.html).

## How to extract parts of Dates

Here we can use the function `format()`. Let say, we want to know the week number of my birthday, we can do:

```{r eval=F}
juan_bh <- as.Date("1979/02/21")
format(juan_bh, format = "Week was %V")    # [1] "Week was 08"
```
We can also check other date elements, like:

```{r eval=F}
format(juan_bh, format = "That day it was %A (%a)")        # [1] "That day it was Wednesday (Wed)"
format(juan_bh, format = "It was the day %u of the week")  # [1] "It was the day 3rd of the week 08"
format(juan_bh, format = "It was %Cth century")            # [1] "It was 19th century"
```



