---
title: "How to deal with Dates and Times in R"
author: "Juan Lorenzo Arellano (@jlarellanoh)"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

A common need that exists in every programming language is parsing strings into dates and converting dates into string. This is specifically important for data analysts because we get data from many different sources and countries where the format changes, and we need to read that data into a R as dates. For instance, 31st of January of 2021 can be seen as:

* **01/31/21**: US format
* **January the 31, 2021**: US format
* **31/01/21**: typically European format
* **31.01.21**: typically German format
* **31-01-21**: typically English format
* **31/JAN/21**: variation of English format
* **31 January 2021**: variation of English format

Conversely, in many occasions we to extract specific parts of the date like day of week, calendar week, month name, day name, etc. or produce a long string containing date parts like "Today is Monday of week 23 (2021)" when creating reports or dashboards.

We are going to see how we can do this transformations in **base R** (we will not cover [**lubridate**](https://lubridate.tidyverse.org/) nor the new kid on the block: [**clock**](https://clock.r-lib.org/)

## How to parse a Character String into a Date

A date is data type or class that exists within R. When you need to write a date into a report, document, csv, etc. you covert into a string. Also, to enter a date into the R terminal you produce a string that has to be converted into an R date type.

```{r}
mydate_chr <- "2021-04-15"
class(mydate_chr)
mydate_date <- as.Date(mydate_chr)  # we do not need to specify any format as yyyy-mm-dd is a default
mydate_date
class(mydate_date)
```

